---
title: "Covid-19 severity predictor with a Random Forest Model"
author: "Leyre Roca"
date: "Last update: `r format(Sys.time(), '%d %B, %Y')`"      
output:
  html_document:
    theme: cosmo         
    highlight: tango      
    toc: true             
    toc_float: true       
    fig_caption: true
---
<style>
table {
    margin: 0 auto !important;
    width: auto !important;
}
.table caption {
    color: #333333;
    font-weight: bold;
    font-size: 1.2em;
    margin-top: 10px;
    margin-bottom: 10px;
    text-align: left;
    caption-side: top;       
    min-width: 500px;        
    white-space: nowrap;
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, 
  warning = FALSE, 
  message = FALSE,
  fig.align = "center")
```

```{r libraries setup, message=FALSE, warning=FALSE}
library(readxl)
library(ggplot2)
library(factoextra)
library(janitor)
library(dplyr)
library(tidyr)
library(stringr)
library(readr)
library(lubridate)
library(patchwork)
library(recipes)
library(randomForest)
library(pROC)
library(tidymodels)
library(vip)
library(ranger)
library(knitr)
library(tune)
```

# Part 1

**Building a Random Forest model to classify COVID-19 severity**

Using the training cohort provided in Supplementary Table 3, a Random Forest classification model is built, to distinguish between non-Severe and Severe COVID-19 patients.

The following performance metrics from cross-validation are repoprted:

-   Confusion matrix\
-   Accuracy\
-   ROC curve and AUC

```{r Preparing metadata, message=FALSE, warning=FALSE}
#Preparing metadata
df_metadata <- read_excel("data/raw_data/Table-S1.Additional-Demographical-and-Baseline-Characteristics-of-COVID-19-Patients-and-Control-Groups.xlsx", sheet = 2)
df_metadata <- clean_names(df_metadata)

# 2) Build the main Table-1 group column for ALL rows first (critical)
df_metadata <- df_metadata %>%
  mutate(
    group = case_when(
      group_d == 0 ~ "Healthy Control",
      group_d == 1 ~ "Non-COVID-19",
      group_d == 2 ~ "Non-severe",
      group_d == 3 ~ "Severe",
      TRUE ~ NA_character_
    )
  ) %>%
  filter(!is.na(group))  # removes any unexpected codes

# 3) Clean variables used in Table 1
df_metadata <- df_metadata %>%
  mutate(
    # sex coding (you confirmed: 1=Male, 0=Female)
    sex = case_when(
      sex_g == 1 ~ "Male",
      sex_g == 0 ~ "Female",
      TRUE ~ NA_character_
    ) %>% factor(levels = c("Male","Female")),

    # numeric age
    age = parse_number(as.character(age_year)))
```    

```{r preparing data for Random Forest, message=FALSE, warning=FALSE}
df_protein <- read_excel("data/raw_data/Table-S3.Protein-Metabolite-Matrix.xlsx", sheet = 2, skip = 1) |>
  rename("Proteins/Metabolites" = 1, "Gene Symbol" = 2) |>  
  clean_names()
df_protein <- df_protein |> filter(gene_symbol != "NA")

#Trasposing data (rows: patients / columns: proteins)
m_t_protein <- t(df_protein[3:ncol(df_protein)]) 
colnames(m_t_protein) <- df_protein$proteins_metabolites

#Random Forest needs data in a dataframe (y que los IDs no sean una columna de variables)
df_t_protein <- as.data.frame(m_t_protein)
df_t_protein <- data.frame(patient_id_a = str_to_upper(rownames(m_t_protein)), df_t_protein)

#Adding Level of Severity
df_2_protein <- df_t_protein |>
  inner_join(df_metadata |> select(patient_id_a, group), by = "patient_id_a")
#Removing IDs
df_2_protein <- data.frame(diagnosis = df_2_protein$group, df_2_protein[3:(ncol(df_2_protein)-1)]) |> mutate(diagnosis = as.factor(diagnosis))
df_2_protein$diagnosis <- factor(df_2_protein$diagnosis, levels = c("Severe","Non-severe"), ordered = TRUE)

```


```{r Random Forest, message=FALSE, warning=FALSE}
#1. Split in Training (80%) and Test (20%) Data
set.seed(123)

#Initial split
data_split <- initial_split(df_2_protein, prop = 0.80, strata = diagnosis)
train_data <- training(data_split)
test_data <- testing(data_split)


# -------- Random Forest --------------

# Cross validation and tuning
cv_folds <- vfold_cv(train_data, v = 5, strata = diagnosis)

# WORKFLOW 
# Define recipe (normalization and  y cleaning variance)
rf_recipe <- recipe(diagnosis ~ ., data = train_data) %>%
  step_filter_missing(all_predictors(), threshold = 0.4) %>% 
  step_impute_knn(all_numeric_predictors(), neighbors = 5) %>% #impute NAs
  step_zv(all_predictors()) %>%
  step_normalize(all_numeric_predictors())

# Define the model with tuning
rf_spec <- rand_forest(trees = 500, mtry = tune() ) %>%
  set_engine("ranger", importance = "impurity") %>%
  set_mode("classification")

# Joing everything in the Workflow
rf_workflow <- workflow() %>%
  add_recipe(rf_recipe) %>%
  add_model(rf_spec)

rf_param <- extract_parameter_set_dials(rf_workflow) %>%  
  finalize(train_data)

# Defining mtry
n_cols <- 500
sqrt_val <- floor(sqrt(n_cols))

mis_valores_mtry <- c(  floor(sqrt_val / 2), sqrt_val, sqrt_val * 2, 
  floor(n_cols / 3), floor(n_cols / 2))
mis_valores_mtry <- unique(pmin(mis_valores_mtry, n_cols))
mtry_grid <- expand.grid(mtry = mis_valores_mtry)

# Execute tuning
rf_results <- tune_grid(
  rf_workflow,
  resamples = cv_folds,
  grid = mtry_grid,
  metrics = metric_set(accuracy, roc_auc, yardstick::sensitivity, yardstick::specificity)
)

autoplot(rf_results)

# Last step: last fit
# Choosing best model based on ROC AUC
best_params <- select_best(rf_results, metric = "roc_auc")
final_wf <- finalize_workflow(rf_workflow, best_params)
my_metrics <- metric_set(accuracy, roc_auc, yardstick::sensitivity, yardstick::specificity)
final_split <- make_splits(train_data, test_data) 
final_res <- last_fit(final_wf, split = final_split,
                      metrics = my_metrics)

#Metrics Results
metrics_original <- collect_metrics(final_res)
kable(metrics_original, caption ="Metrics Results")


# Confusion Matrix
collect_predictions(final_res) %>%
  conf_mat(truth = diagnosis, estimate = .pred_class) %>%
  pluck("table") %>% # Extraemos solo la matriz numérica
  kable(caption = "Confusion Matrix")

###### Lowering classification threshold from 0.5 to 0.45
preds <- collect_predictions(final_res)

# Preparing dataframe with the new class with the 0.45 threshold
preds_ajustadas <- collect_predictions(final_res) %>%
  mutate(
    .pred_class_045 = if_else(.pred_Severe > 0.45, "Severe", "Non-severe"),
    .pred_class_045 = factor(.pred_class_045, levels = c("Severe", "Non-severe"))
  )

# Confusion matrix
preds_ajustadas %>% 
  conf_mat(truth = diagnosis, estimate = .pred_class_045)  %>%
   pluck("table") %>% 
   kable(caption = "Final Confusion Matrix")


# Metrics Selection
mis_metricas <- metric_set(accuracy, yardstick::sensitivity, yardstick::specificity)
tabla_metricas_clase <- preds_ajustadas %>%
  mis_metricas(truth = diagnosis, estimate = .pred_class_045)

# 4. Calculate metrics probabilities 
metricas_prob <- collect_predictions(final_res) %>%
  metric_set(roc_auc)(truth = diagnosis, .pred_Severe)

# Join result
resultado_final <- bind_rows(tabla_metricas_clase, metricas_prob)
kable(resultado_final, caption = "Final Metrics Results")

# ROC Curve
collect_predictions(final_res) %>%
  roc_curve(truth = diagnosis, .pred_Severe) %>% 
  autoplot()
```

**Evaluation of how well does the Random Forest model separate non-Severe from Severe patients based on proteomic profiles**

Initially, the random forest model had trouble identifying Severe patients, yielding a very low sensitivity of 0.33. In order to improve the detection of Severe patients, the classification threshold was adjusted from 0.50 to 0.45. 

After that adjustment, the sensitivity of the model improved from 0.33 to 0.66. And the accuracy value of the model improved from 0.57 to 0.71, while maintaining the AUC value at 0.83 and the specifity of the model at 0.75.

With this results, the model it's better than a random model but still it isn't sufficiently robust for classifying Severe COVID patients. This can be partly explained by small sample size (n=26 patients) used for the training and the fact that the imputing of the data had to be adapted to this limited dataset, which may affect the model's generalizability.


# Task 2

**Identifying and interpreting the most important features**

Using the trained Random Forest model from Exercise 1, the script identifies the 25 most important protein features contributing to the classification. Their biological relevance is discussed.


```{r, message=FALSE, warning=FALSE}
df_S5 <- read_excel("data/raw_data/Table-S5.Protein-Metabolite-Matrix.xlsx", sheet = 2, skip = 1) |>
  rename("Proteins/Metabolites" = 1, "Gene Symbol" = 2) |>  clean_names() |> filter(gene_symbol != "NA")

# Extract trained model from object final_res
final_fit_model <- final_res %>% 
  extract_fit_parsnip()

# Importance feature plot
final_fit_model %>%
  vip(num_features = 25, geom = "point") +
  theme_minimal() +
  labs(title = "Top 25 Proteins")

# Save important_features results as a data frame
table_importance <- final_fit_model %>%
  vi()
top25_proteins <- data.frame(proteins_metabolites = table_importance$Variable, importance = table_importance$Importance)

print("The top 25 most important protein features for my model are the following:")
top25_proteins$proteins_metabolites[1:25]

protein_match <- c()
for (protein in df_S5$proteins_metabolites){
  if (any(grepl(protein, top25_proteins$proteins_metabolites[1:25], fixed = TRUE))) {
    protein_match <- append(protein_match, protein)
  }
}

print("The common proteins found bwtween my top 25 most important protein features for my model and the proteins found in Supplementary Table 5 are the following:")
protein_match
```

**Discussion of the biological relevance of the proteins.**

The selection is dominated by acute-phase reactants (SAA1, SAA2, ORM1) and components of the complement and coagulation cascades, reflecting the systemic inflammation response characteristic of severe cases. The inclusion of P0DJI8 (SAA1) and P0DJI9 (SAA2) is particularly significant, as these markers are established indicators of COVID-19 progression from non-severe to severe states.


**Comparison of the selected features to the proteins reported by the authors in Supplementary Table 5.**

Comparing my top 25 features with the proteins found in Supplementary Table 5, we can observe several common markers, as biomarkers of extreme inflammation (P0DJI8 (SAA1) and P0DJI9 (SAA2)), endothelial dysfunction and tissue damage (P02750 (LRG1) and P05452 (Tetranectin) and negative acute phase reactants (P02768 (ALB)). Finding this overlap of highly relevant COVID-19 markers indicates that my model captures the essence of the COVID-19 proteomic signature reported in the literature.


# Part 3

**External prediction on an independent test cohort**

Using the final Random Forest model trained on the full training cohort, disease severity is predicted for patients in the independent test cohort provided in Supplementary Table 4.

Objectives of this block:

-   Applying the trained model to the test cohort.  
-   Obtaining predicted probabilities of Severe disease for each patient.  
-   Assigning predicted class labels based on these probabilities.  
-   Comparing predictions to the true clinical labels.  


```{r Preparing S4 data, message=FALSE, warning=FALSE}
df_external <- read_excel("data/raw_data/Table-S4.Protein-Metabolite-Matrix-C2.xlsx", sheet = 2, skip = 1) |>
  rename("Proteins/Metabolites" = 1, "Gene Symbol" = 2) |>  
  clean_names() |> filter(gene_symbol != "NA")

# Preparing data 
m_t_external <- t(df_external[3:ncol(df_external)]) 
colnames(m_t_external) <- df_external$proteins_metabolites
df_t_external <- as.data.frame(m_t_external)
df_t_external <- data.frame(patient_id_a = str_to_upper(rownames(m_t_external)), df_t_external)

# Adding Level of Severity
df_2_external <- df_t_external |>
  inner_join(df_metadata |> select(patient_id_a, group), by = "patient_id_a")
# Removing IDs
df_2_external <- data.frame(diagnosis = df_2_external$group, df_2_external[3:(ncol(df_2_external)-1)]) |> mutate(diagnosis = as.factor(diagnosis))
df_2_external$diagnosis <- factor(df_2_external$diagnosis, levels = c("Severe","Non-severe"), ordered = TRUE)


proteinas_ausentes <- setdiff(colnames(df_2_protein), colnames(df_2_external))
criticas <- intersect(proteinas_ausentes, top25_proteins$gene_symbol[1:25])

if (length(criticas) < 1) {
  print("Missing proteins in Supplementary Table S4. Retraining Random Forest model with only the common proteins.")
}

```

```{r external prediction on Cohort C2, message=FALSE, warning=FALSE}

### RETRAINING THE MODEL ###
set.seed(123)

# Both datasets have the same columns
common_cols <- intersect(colnames(df_2_protein), colnames(df_2_external))
train_data_S4 <- df_2_protein |> select(all_of(common_cols))
test_data_S4 <- df_2_external |> select(all_of(common_cols))

# WORKFLOW 
# Define recipe (normalization and  y cleaning variance)
rf_recipe_S4 <- recipe(diagnosis ~ ., data = train_data_S4) %>%
  step_filter_missing(all_predictors(), threshold = 0.3) %>% 
  step_impute_knn(all_numeric_predictors(), neighbors = 5) %>% #imputar NAs
  step_zv(all_predictors()) %>%
  step_normalize(all_numeric_predictors())

# Define the model with tuning
rf_spec_S4 <- rand_forest(
    trees = 500,
    mtry = tune() 
  ) %>%
  set_engine("ranger", importance = "impurity") %>%
  set_mode("classification")

# Join everything in the Workflow
rf_workflow_S4 <- workflow() %>%
  add_recipe(rf_recipe_S4) %>%
  add_model(rf_spec_S4)

# Cross-validation and tuning
cv_folds_S4 <- vfold_cv(train_data_S4, v = 5, strata = diagnosis)

# setting mtry
rf_param <- extract_parameter_set_dials(rf_workflow_S4) %>%  
  finalize(train_data_S4)
n_cols <- 500
sqrt_val <- floor(sqrt(n_cols))
mis_valores_mtry <- c(  floor(sqrt_val / 2), sqrt_val, sqrt_val * 2, 
  floor(n_cols / 3), floor(n_cols / 2))
mis_valores_mtry <- unique(pmin(mis_valores_mtry, n_cols))
mtry_grid <- expand.grid(mtry = mis_valores_mtry)

# Execute tuning
rf_results_S4 <- tune_grid(
  rf_workflow_S4,
  resamples = cv_folds_S4,
  grid = mtry_grid,
  metrics = metric_set(accuracy, roc_auc, yardstick::sensitivity, yardstick::specificity)
)

autoplot(rf_results_S4)

# Last Step : Last Fit
# Elegimos el mejor modelo basado en ROC AUC
best_params_S4 <- select_best(rf_results_S4, metric = "roc_auc")
final_wf_S4 <- finalize_workflow(rf_workflow_S4, best_params_S4)
my_metrics_S4 <- metric_set(accuracy, roc_auc, yardstick::sensitivity, yardstick::specificity)
final_split_S4 <- make_splits(train_data_S4, test_data_S4) 
final_res_S4 <- last_fit(final_wf_S4, split = final_split_S4,
                      metrics = my_metrics_S4)

# Metrics Results
metrics_original <- collect_metrics(final_res_S4)
kable(metrics_original, caption = "Metrics Results")

# Confusion Matrix
collect_predictions(final_res_S4) %>%
  conf_mat(truth = diagnosis, estimate = .pred_class) %>%
  pluck("table") %>% # Extraemos solo la matriz numérica
  kable(caption = "Confusion Matriz")

###### Lowering classification threshold from 0.5 to 0.45

preds <- collect_predictions(final_res)

# Prepare dataframe with the new class using threshold 0.45
preds_ajustadas_S4 <- collect_predictions(final_res_S4) %>%
  mutate(
    .pred_class_045 = if_else(.pred_Severe > 0.45, "Severe", "Non-severe"),
    .pred_class_045 = factor(.pred_class_045, levels = c("Severe", "Non-severe"))
  )

# Confusion Matrix
preds_ajustadas_S4 %>% 
  conf_mat(truth = diagnosis, estimate = .pred_class_045)  %>%
   pluck("table") %>% 
   kable(caption = "Final Confusion Matrix")

# Define metrics selection 
mis_metricas <- metric_set(accuracy, yardstick::sensitivity, yardstick::specificity)

# Calculate metrics (for threshold 0.45)
tabla_metricas_clase_S4 <- preds_ajustadas_S4 %>%
  mis_metricas(truth = diagnosis, estimate = .pred_class_045)
metricas_prob_S4 <- collect_predictions(final_res_S4) %>%
  metric_set(roc_auc)(truth = diagnosis, .pred_Severe)

# Joining the results
resultado_final_S4 <- bind_rows(tabla_metricas_clase_S4, metricas_prob_S4)
kable(resultado_final_S4, caption = "Final Metrics Results")

#ROC Curve
collect_predictions(final_res_S4) %>%
  roc_curve(truth = diagnosis, .pred_Severe) %>% # Ajusta 'Severe' a tu clase positiva
  autoplot()
```

**Evaluation of the model performance on the independent test cohort**

For assigning the class labels to patients from the independent cohort (n=10), the model with a classification threshold adjusted from 0.50 to 0.45 was applied. The accuracy of this model in the independent cohort was 0.70 (improving from 0.60 in the original model), with a sensitivity of 1.0 (improved from 0.75), a specificity of 0.50 and an AUC value of 0.83. 

With these results, we can conclude that while the model perfectly identifies Severe patients (zero false negatives), its ability for identifying Non-severe patients is weak, as it only correctly predicts half of them, leading to a high number of false positives (Non-severe patients classified as Severe). For these reasons, the current model could be useful as a high-sensitivity screening tool but it wouldn't be suitable for a definitive clinical diagnosis.

# Session info {.unnumbered}
<details>
<summary><strong>Session Information</strong> (Click to show)</summary>

```{r, results='asis',  echo=FALSE, message=FALSE }
sessionInfo()
```
